---
title: "FootprintAnalysis"
author: "Laura van der Poel"
date: "`r Sys.Date()`"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/l_vdp/Documents/MSc_Internship/tower_fps')
```

This R Markdown document is the accompanying script for the Research Report: "The effect of tree cover on daytime carbon fluxes of a Mediterranean oak woodland" by Laura van der Poel. For details about this study, consult the report. 

Here, all the necessary steps for the footprint calculation based on the model by Kljun et al. (2015) are shown. These footprints are overlaid with the tree shapes, and tree cover is calculated.

```{r message=FALSE} 
# load the libraries
rm(list=ls())

library(stringr)
library(tidyterra)
library(sf)
library(tidyr)
library(sp)
library(beepr)
library(lubridate) 
library(EBImage)
library(progress)
library(terra)
library(ncdf4)
library(flux)
library(FME)


```

#### Create dataframe for FP calculation 

The data is the raw data from the central eddy covariance tower in Majadas, as preprocessed by the Max Planck Institute.

```{r load-data}
# load Flux data
FPathF        <- "./data/Majadas_main_2015-20.csv"
dfF_all.na    <- read.csv(FPathF, row.names=1) 
dfF_all.na$rDate <- as_datetime(dfF_all.na$rDate)  

# delete rows with NA's in necessary columns for footprint model
dfF_all           <- dfF_all.na %>% drop_na(v_sd, rDate, umean, wind_dir, ustar, L) 

```

#### Get Boundary Layer Height

The boundary layer height (blh) was downloaded from https://cds.climate.copernicus.eu/cdsapp#!/software/app-era5-explorer?tab=overview, in netcdf format.

```{r get-blh}
# get data
ncpath      <- "./data/ERA5/adaptor.mars.internal-2015-2020.nc"
nc_data     <- nc_open(ncpath)

# extract blh, latitude, longitude and time data
blh_array   <- ncvar_get(nc_data, "blh")
lat         <- ncvar_get(nc_data, "latitude")
lon         <- ncvar_get(nc_data, "longitude")
time        <- ncvar_get(nc_data, "time") # unit of time is "hours since 1900-01-01 00:00:00.0"

# transform hours from 1900-01-01 to date 
time_obs    <- as.POSIXct(time*3600, origin = "1900-01-01" )

# get FillValue and replace with NA
fillvalue   <- ncatt_get(nc_data, "blh", "_FillValue")
blh_array[blh_array==fillvalue$value] <- NA

# Create 2D matrix of long, lat and time
lonlattime  <- as.matrix(expand.grid(lon,lat,time_obs))

# reshape whole blh_array
blh_vec_long <- as.vector(blh_array)

# create data.frame
blh_obs    <- data.frame(cbind(lonlattime, blh_vec_long))
colnames(blh_obs) <- c('lon', 'lat', 'timestamp', 'blh') 

# linear interpolation to 30 minutes
blh_obs$timestamp <- ymd_hms(blh_obs$timestamp)
blh_li            <- lips(blh_obs$timestamp, as.numeric(blh_obs$blh), x.step = 1800)
colnames(blh_li)  <- c('rDate', 'blh')


# join dataframes
dfF_all       <- left_join(dfF_all, blh_li, by='rDate')

```

#### Footprint preparation

We filter on u-star threshold, calculated as in Papale et al. (2006) by the Max Planck Institute. 

```{r filter-ustar}
# ustar threshold
ustar_thres <- unique(dfF_all$Ustar_uStar_Thres)[3] # 0.21
dfF_all     <- dfF_all[dfF_all$ustar > ustar_thres, ] 

```

Other necessary information for the footprint function

```{r settings}

# settings for footprint model

Fzm           <- 15   # Flux measurement height
CHeight       <- 8    # Canopy height
DomX          <- 500  # FP Domain settings
DomY          <- 500  # FP Domain settings


dfF_all$zmd   <- Fzm-(2/3)*CHeight # Measurement height above displacement height, needed in the model
```

Select begin and end date of analysis.

```{r beg-end-date}
# crop df till Beg-EndDate  
Title         <- 'Jan_2015' # title will be later used to store results
BegDate       <- ymd_hm("2015-01-01 00:00")    
EndDate       <- ymd_hm("2015-02-01 00:00")     

dfF   <- dfF_all[dfF_all$rDate >= BegDate & dfF_all$rDate < EndDate,
                 # variables for FP calculation
                   c("rDate","umean","wind_dir","ustar","L","v_sd","zmd","blh", "day_night")]

```

#### Optimize the roughness length per wind direction bin

Calculate z0 for the selected subset of data. 

```{r optimize-z0, results = FALSE, message=FALSE}

# function to add column to dataframe, conatining wind direction bin 
add_wind_dir_bins <- function(df, degrees_per_bin=30){
  
  # create wind direction bins of 30 degrees
  bins <- seq(0,360,degrees_per_bin)
  
  # group dataframe
  df <- df %>% mutate(wind_dir_bins = cut(wind_dir, breaks=bins))
  
  # give correct names
  wind_dirs <- c('NNE','NE', 'ENE', 'ESE','SE', 'SSE', 'SSW', 'SW', 'WSW', 'WNW', 'NW',  'NNW')
  levels(df$wind_dir_bins) <- wind_dirs
  
  return(df)}

# this function filters the dataframe to obtain only measurements of stable conditions
subset_for_z0_calc <- function(df){
  df.ss <- df[(abs(Fzm / df$L)) < 0.1 & 
                df$ustar > 0.21 &
                df$ustar < 0.8 &
                !is.na(df$wind_dir_bins),]
  return(df.ss)}

# formula for Uz that will be optimized
u_fun_z0 <- function(par, ustar, umean, k=0.4, z=15, d=9.666667){
  z0 = par
  
  # function for Uz
  Uz <- (ustar / k) * log((z - d)/ z0)
  
  # residuals
  res <- umean - Uz 
  return(res)
}

# function that finds optimized z0 for specific wind direction bin
opt_z0 <- function(wind_dir, df.ss, p = 0.8){
  
  df.wind_dir <- df.ss[df.ss$wind_dir_bin == wind_dir, ]
  ustar <- df.wind_dir$ustar
  umean <- df.wind_dir$umean
  
  fit <- modFit(f=u_fun_z0, p=0.8, lower=0, upper=20, ustar = ustar, 
                umean = umean)
  
  summ <- summary(fit)
  modvar <- summ$modVariance
  
  mcmc <- modMCMC(f=u_fun_z0, p=0.8, ustar=ustar, umean=umean, jump = c(0.05, 0.05), 
                  lower = 0, upper = 20, prior =NULL, var0=modvar, niter=5000, outputlength = 1000)
  
  return(mcmc$bestpar)
}


# get results of optimizer function
get_results <- function(df.ss){
  wind_dirs <- c('NNE','NE', 'ENE', 'ESE','SE', 'SSE', 'SSW', 'SW', 'WSW', 'WNW', 'NW',  'NNW')
  # create dataframe to store results
  results <- data.frame(wind_dir_bins = wind_dirs, z0 = NA)
  
  for(wind_dir in wind_dirs){
    
    tryCatch({
    bestpar <- opt_z0(wind_dir, df.ss=df.ss)
    results[results$wind_dir_bins == wind_dir, 'z0'] <- bestpar
    
  }, error=function(e){})
    
  }
  return(results)
}

# now apply functions to our data
dfF          <- add_wind_dir_bins(dfF) 
dfF.ss       <- subset_for_z0_calc(dfF)

z0_wind_dirs <- get_results(dfF.ss)

```

#### Load spatial data

Now, load the spatial data: towers, tree locations, and tree areas
The tower coordinates are taken from the Towers.shp file, and stored in a terra-object. We will use this package for all the spatial data and calculations. 

```{r}
# store tower coordinates in correct format and crs
TowerXY       <- matrix(c(2976558.03329999931156635,2012190.60920000076293945), ncol=2) 
TowerXY       <- vect(TowerXY, type="points", crs="EPSG:3035")
TowerXY       <- project(TowerXY, 'EPSG:32629')

```

Load the tree data, both locations of trees as areas

```{r}
# area trees
treeArea.path  <- "./data/copas_cir_maj_2019/copas_cir_maj_2019.shp"
treeArea       <- vect(treeArea.path)

# location trees
treeLoca.path  <- "./data/Trees_Majadas/Trees_Majadas.shp"
treeLoca       <- vect(treeLoca.path) # crs = 'EPSG:3035'
treeLoca_p     <- project(treeLoca,  'EPSG:32629')
```


### Create functions to overlay footprints with trees

These functions are used in analysing FPOut, the result of the footprint model.

This function takes the values of the footprint distribution [m-2], and shifts them to the tower location in a SpatRaster
```{r get-fp-distr}
get_fp_distr <- function(FPOut, DomX, DomY, TowerXY){
    
    
    fp_values                <- FPOut$fclim_2d
    fp_values[fp_values < 0] <- 0
    fp_values                <- fp_values / sum(fp_values)

    # fp values as SpatRaster, and shift to location of tower
    fp_rast      <- raster::raster(t(fp_values[,ncol(fp_values):1]),
                                    xmn = -DomX,
                                    xmx = DomX,
                                    ymn = -DomY,
                                    ymx = DomY,
                                    crs = 32629) %>% rast()

    fp_rast_sh    <- shift(fp_rast, dx=crds(TowerXY)[1], dy=crds(TowerXY)[2])
    
    return(fp_rast_sh)
    }
```


This function gets the first (80%) contour line of the footprint, and stores it in a SpatVector.

```{r get-fp-cont}
get_fp_cont <- function(FPOut, TowerXY){
  
  # store xr & yr (the x and y array from the contour lines) as a SpatVector
  contour_vect <- cbind(FPOut$xr[[1]], as.array(FPOut$yr)[[1]]) %>% 
    vect() %>% as.lines()
  
  # shift to tower location
  contour_vect_sh <- shift(contour_vect, dx=crds(TowerXY)[1], dy=crds(TowerXY)[2])
  crs(contour_vect_sh) <- 'EPSG:32629'
  return(contour_vect_sh)
}
```

This function calculates the tree density and the tree canopy fraction in the footprint contour.

```{r calc-trees-fp}
calc_trees_fp <- function(contour_vect_sh, treeArea.path,
                          treeLoca_p){
  # crop data to contour 
  
  contour_poly       <- as.polygons(contour_vect_sh)
  treeArea_c         <- vect(treeArea.path, filter= as.polygons(ext(contour_vect_sh))) # crs = 'EPSG:32629'
  treeLoca_c         <- crop(treeLoca_p, ext(contour_vect_sh))
  
  # store area in footprint contour
  
  FPArea      <- expanse(contour_poly, unit='m')
  
  # 1: count number of trees in contour
  
  n_trs         <- terra::extract(contour_poly, treeLoca_c) %>% data.frame()
  TreeCount     <- sum(n_trs$id.x, na.rm=TRUE)   # number of trees in footprint
  TreeDens      <- (TreeCount  / FPArea) * 10000    # density of trees: number of trees per hectare
  
  #2: sum area of trees in contour
  
  a             <- terra::intersect(contour_poly, treeArea_c)
  TreeArea      <- sum(expanse(a, unit='m'))  # area of trees in footprint
  TreeArea_frac <- TreeArea / FPArea   # fraction of area of trees in footprint
  
  return(list=c(FPArea=FPArea, TreeCount=TreeCount, TreeDens=TreeDens, 
                TreeArea=TreeArea, TreeArea_frac=TreeArea_frac))

}
```

Initialize progress par from {progress} package

```{r}
n_iter = nrow(dfF) * 2
pb <- progress_bar$new(format = "(:spin) [:bar] :percent [Elapsed time: :elapsedfull || Estimated time remaining: :eta]",
                       total = n_iter,
                       complete = "=",   # Completion bar character
                       incomplete = "-", # Incomplete bar character
                       current = ">",    # Current bar character
                       clear = FALSE,    # If TRUE, clears the bar when finish
                       width = 100)      # Width of the progress bar
```

#### Footprint model 

Finally, get the footprint model itself 

```{r fp-preparation}
# load the function 
source("./scripts/functions_kljun/calc_footprint_FFP_climatology.R")

# path to store footprint
SPath         <- "./footprints/"

```

#### Loop over entire dataframe

Loop over the dataframe, to calculate the footprint and the tree cover and density for every observation. An extra loop is built in to try both z0=NA, and z0=the optimised z0 per wind direction bin. 


```{r for-loop, results = FALSE, message=FALSE}

StartTime     <- Sys.time()

for (i in 1:nrow(dfF)){
  
  # loop over z0-types, Kljun 2015 and Pennypacker Baldocchi 2015
  for(z0_type in c('K15', 'PB15')){
    if(z0_type == 'K15'){z0 = NA}
    if(z0_type == 'PB15'){z0 = z0_wind_dirs[z0_wind_dirs$wind_dir_bins == 
                                              dfF[i, 'wind_dir_bins'], 'z0']}
  tryCatch({
    print(paste0(i, '/', nrow(dfF)))
    # progress bar, updates the current state
    pb$tick()
    
    #### run FP ####
    FPOut       <- calc_footprint_FFP_climatology(zm        = dfF$zmd[i],
                                                  z0        = z0,
                                                  umean     = dfF$umean[i],
                                                  h         = dfF$blh[i],
                                                  ol        = dfF$L[i],
                                                  sigmav    = dfF$v_sd[i],
                                                  ustar     = dfF$ustar[i],
                                                  wind_dir  = dfF$wind_dir[i],
                                                  domain    = c(-DomX,DomX,-DomY,DomY),
                                                  dx = 1,dy = 1,
                                                  r         = 0.8,
                                                  crop      = 0,
                                                  rslayer   = 1 
                                                  # fig = 1 (can be added to produce figure)
                                                  )
    FPTimeStamp  <- paste0(year(dfF$rDate[i]),
                          str_pad(string = month(dfF$rDate[i]),width = 2,side = 'left',pad = '0'),
                          str_pad(string = mday(dfF$rDate[i]),width = 2,side = 'left',pad = '0'),
                          str_pad(string = hour(dfF$rDate[i]),width = 2,side = 'left',pad = '0'),
                          str_pad(string = minute(dfF$rDate[i]),width = 2,side = 'left',pad = '0'))
    
    # fp values in a raster, shifted to tower location 
    fp_rast_sh <- get_fp_distr(FPOut=FPOut, DomX=DomX, DomY=DomY, TowerXY=TowerXY)
    
    # footprint contourline (80%) as SpatVector, and shift to location of tower
    contour_vect_sh <- get_fp_cont(FPOut=FPOut, TowerXY = TowerXY)

    # calculate trees in footprint 
    tree_values <- calc_trees_fp(contour_vect_sh=contour_vect_sh, 
                              treeArea.path=treeArea.path, treeLoca_p=treeLoca_p)
    
    # save in dataframe with added title for the roughness length that was used
    names(tree_values) <- lapply(names(tree_values), function(x) paste0(x, '_',z0_type))
    
    dfF[i, names(tree_values)] <- tree_values
  
    }, warning = function (war) {
    print(paste0("Footprint ", i, " did not work well, because ", warning))
  }, error = function (err) {
    print(paste0("Footprint ", i, " did not work, because ", err))
  })
  }
}

EndTime <- Sys.time()
print(EndTime-StartTime)
```
 
And store results in a .csv file


```{r}
write.csv(dfF, paste0(SPath, Title ,  '_FPdata_trees.csv'))

```

In the research, also the footprints and the contour lines were stored and analysed. This part was left out of the current script, but can be requested if needed.
